---
title: "Plots and exploratory data analysis in R"
subtitle: "Data analysis and visualization in R <br> UC Merced"
author: "Andrea Sánchez-Tapia - Sara Ribeiro Mortara<br> ¡liibre! - RLadies+"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: true
      ratio: '16:9'
---
## last time

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(htmltools.dir.version = FALSE)
library(knitr)
```

```{r xaringanthemer, include = FALSE}
xaringanExtra::use_share_again()
```

+ we talked about __matrices__ and __lists__ using function `matrix()` as an example
+ we talked about data frame objects, `str()`, `dim()`, `nrow()`, `ncol()`, and subsetting `[rows, columns]`
+ we downloaded a file, read it into disk, removed rows with NAs and saved it back into a __processed__ data folder
+ we talked about __factors__: in R>4.0 you need to specify them with `factor()`

## today

+ exploratory data analysis [__Why__ do we plot our data?]
+ basic plotting functions [__How__ do we plot our data?]

---
class: inverse, middle, center

# Exploratory data analysis

---
## exploratory data analysis (EDA)

+ control the quality of your data

--

+ support the selection of statistical procedures 

--

+ evaluate if data attend the __assumptions__ of the statistical tests (ex. normality)

--

+ suggest hypotheses for the relationship of your data and new studies


--

+ __EDA is NOT data wrangling or manipulation__

--

+ your hypotheses based on theory are __central__ to guide these analyses

---
## exploratory data analysis (EDA)

+ EDA can take 20-50% of your analysis time

--

+ it should be performed during the data collection

--

+ uses a lot of visual techniques

--

+ EDA will help you __understand__ your data

---
## what we need to know about our data 

+ do they contain NAs? do we have a lot of zeroes?

--

+ how are the variables distributed? are they centered? are they symmetric? bimodal? 
  
--

+ are there outliers?

--

+ do the variables follow some distribution?

--

+ do they need to be transformed?

--

+ are the variables related? what is the shape of the relationship between variables? (ex. linear)

--

+ was the sampling effort sufficient? 


---
## what we need to know about our data 

+ central tendency measures: mean, median, mode

--

+ variation/dispersion measures: range, range width, variance, standard deviation, variation coefficient

--

+ data distribution: quantiles, inter-quantile ranges, _boxplots_, _histograms_.

--

+ relationship between variables: _scatterplots_, correlations, linear models

---
class: inverse, center, middle

## The Anscombe quartet

---
## The Anscombe quartet

```{r}
data("anscombe")
dim(anscombe)
head(anscombe)
```

---
## The Anscombe quartet

```{r}
class(anscombe)
str(anscombe)
```

---
## Central tendency measures

```{r}
mean(anscombe$x1)
mean(anscombe$x2)
mean(anscombe$x3)
mean(anscombe$x4)
```

---
## Central tendency measures

```{r}
apply(anscombe[,1:4], 2, mean)
apply(anscombe[,5:8], 2, mean)
apply(anscombe, 2, var)

```

---
## Correlations

```{r}
cor(anscombe$x1, anscombe$y1)
cor(anscombe$x2, anscombe$y2)
cor(anscombe$x3, anscombe$y3)
cor(anscombe$x4, anscombe$y4)
```

---
## Linear regression parameters

+ remember a linear regression: __y = a + bx__, where a is the intercept and b is the slope

```{r, eval = T}
m1 <- lm(anscombe$y1 ~ anscombe$x1)
m2 <- lm(anscombe$y2 ~ anscombe$x2)
m3 <- lm(anscombe$y3 ~ anscombe$x3)
m4 <- lm(anscombe$y4 ~ anscombe$x4)

coef(m1)
coef(m2)
coef(m3)
coef(m4)
```

---
## Linear regression coefficients

```{r}
mlist <- list(m1, m2, m3, m4)
lapply(mlist, coef)
```

---
## Let's plot the Anscombe data

```{r, eval = F, echo = TRUE, fig.width=1200, fig.height = 400}
#par(mfrow = c(2,2),
#    las = 1,
#    bty = "l")
plot(anscombe$y1 ~ anscombe$x1)
abline(mlist[[1]])
plot(anscombe$y2 ~ anscombe$x2)
abline(mlist[[2]])
plot(anscombe$y3 ~ anscombe$x3)
abline(mlist[[3]])
plot(anscombe$y4 ~ anscombe$x4)
abline(mlist[[4]])
#par(mfrow=c(1, 1))
```

---
## one example EDA workflow


```{r}
data(iris)
#head(iris)
summary(iris)
```

---
## how many observations do we have?

```{r, eval = F}
table(iris$Species)
plot(iris$Species) #barplot is the default funciton when you plot a categorical variable
```

---
## central tendency measures

```{r, eval = F}
mean(iris$Sepal.Length)
median(iris$Sepal.Length)

## for each species: 
tapply(X = iris$Sepal.Length,
       INDEX = iris$Species,
       FUN = mean)

tapply(X = iris$Sepal.Length,
       INDEX = iris$Species,
       FUN = median)
```

---
#central tendency measures

```{r, eval = F}
freqf <- sort(table(iris$Sepal.Length), 
              decreasing = TRUE)
freqf[1] #the most common value (mode) is 5, it appears 10 times
```

---
## data dispersion measures

```{r}
range(iris$Sepal.Length)
diff(range(iris$Sepal.Length))
```

---
## data dispersion measures

+ variance, standard deviation

```{r, eval = F}
var(iris$Petal.Length) # variance
sd(iris$Petal.Length) #standard deviation
sd(iris$Petal.Length) / mean(iris$Petal.Length) * 100 # variation coefficient
```

---
## data dispersion measures

+ for each species?

```{r, eval = F}
tapply(X = iris$Sepal.Length, INDEX = iris$Species, FUN = sd)
tapply(X = iris$Sepal.Width, INDEX = iris$Species, FUN = sd)
```


---
## data distribution: quantiles and inter-quantile range (IQR)

```{r qq}
quantile(iris$Petal.Length) #quantiles
quantile(iris$Petal.Length, probs = c(0.05, 0.5, 0.95)) #other quantiles
IQR(iris$Petal.Length) #inter-quantile range
summary(iris$Petal.Length)
```


---
## data distribution: boxplot

.pull-left[
```{r, fig.align='left', echo=TRUE, out.width=450}
boxplot(iris$Petal.Length)
```
  ]

.pull-right[  
```{r, echo=FALSE,  fig.align='right', out.width=450}
set.seed(2)
par(bty = "n")
bp <- rnorm(1000, 0, 0.1)
boxplot(bp, yaxt = "n", xlim = c(0, 3), ylim=c(-0.3, 0.3))
text(x = 2, y = min(bp), "last point (-1.5 x IIQ)", cex = 2)
text(x = 2, y = quantile(bp)[2], "first quantile", cex = 2)
text(x = 2, y = median(bp), "median", cex = 2)
text(x = 2, y = quantile(bp)[4], "third quantile", cex = 2)
text(x = 2, y = bp[203], "last point (+1.5 x IIQ)", cex = 2)
arrows(x0 = 0.68, x1 = 0.68, y0 = quantile(bp)[2], y1 = quantile(bp)[4], code =
         3, length = 0.05)
text(x = 0.54, y = median(bp), "IIQ", cex = 2)
```
]

---
##  histogram
  
```{r, fig.align='center', echo=TRUE, out.width=400, eval = F}
hist(iris$Sepal.Width)
hist(iris$Sepal.Length)
hist(iris$Petal.Length)
```

---
## histogram types
  
  
```{r, echo=TRUE, fig.width=9, fig.height=4.5}
par(mfrow = c(1,2))
hist(iris$Sepal.Length)
hist(iris$Sepal.Length, probability = TRUE) # empirical probabilistic density curve
par(mfrow = c(1,1))
```

---
## histogram types
  
  
```{r, echo=TRUE, fig.width=9, fig.height=4.5}
par(mfrow = c(1,2))
plot(density(iris$Sepal.Width))
hist(iris$Sepal.Width, probability = TRUE) # empirical probabilistic density curve
lines(density(iris$Sepal.Width), col="blue")
par(mfrow = c(1,1))
```


---
## histogram breaks
  
  
```{r, echo=TRUE, fig.width=9, fig.height=3}
par(mfrow = c(1,3))
hist(iris$Petal.Length)
hist(iris$Petal.Length, 
     breaks = seq(0, max(iris$Petal.Length), length = 4))
hist(iris$Petal.Length,  
     breaks = seq(0, max(iris$Petal.Length), length = 6))
par(mfrow = c(1,1))
```


---
## relationships between variables: scatterplot
  
```{r, eval =T, results = "hide"}
x <- iris$Petal.Length
y <- iris$Petal.Width
plot(x, y) 
#but you can also avoid creating the vectors
plot(iris$Petal.Length, iris$Petal.Width) #check the help for parameter order, x, y axis.
plot(iris$Petal.Width ~ iris$Petal.Length)
```

+ check the __tilde__ `~` notation, formula: "as a function of"

---
## relationship between variables: correlation
  
  
```{r correlation}
cor(x, y)
```

+ when is a correlation high? (~0.7?)

---
## let's go back to out scatterplot

```{r, eval = T, echo = F}
plot(iris$Petal.Length, iris$Petal.Width)

```

---
## plotting basics

+ All parameters for plotting are in function `par()`

+ `pch`, `cex`, `xlab`, `ylab`, `las`

+ `par(mfrow = c(1, 2))`


---
## let's go back to out scatterplot

```{r, eval = F}
plot(iris$Petal.Length, iris$Petal.Width)
plot(iris$Petal.Length, iris$Petal.Width, 
     xlab = "Petal Length (mm)", 
     ylab = "Petal width (mm)", pch = 19)
lmod <- lm(Petal.Width ~ Petal.Length, data = iris)
coef(lmod)
abline(lmod, col = "red")

```
---
## what about species?

```{r, echo = F}
plot(iris$Petal.Length, iris$Petal.Width, 
     xlab = "Petal Length (mm)", 
     ylab = "Petal width (mm)", pch = 19, 
     col = iris$Species)
```

---
## what about species?

```{r, echo = F}
pal <- wesanderson::wes_palette(name = "Darjeeling1", n = 3)
plot(iris$Petal.Length, iris$Petal.Width, 
     xlab = "Petal Length (mm)", 
     ylab = "Petal width (mm)", pch = 19, 
     col = pal[iris$Species])
as.character(pal)
```


---
## let's go back to boxplots

```{r, echo = F}
colors <- c("#3B9AB2", "#EBCC2A", "#F21A00")
sp <- paste("I.", unique(iris$Species), sep = " ")
par(mfrow = c(2, 2),
    mar = c(4, 3, 2, 2), 
    bty = 'l',
    las = 1)
boxplot(Sepal.Length ~ Species,
        data = iris,
        xlab = "",
        col = colors,
        xaxt = "n")
axis(1, at = 1:3, labels = sp, font = 3)
boxplot(Sepal.Width ~ Species, 
        data = iris, 
        xlab = "", 
        col = colors,
        xaxt = "n")
axis(1, at = 1:3, labels = sp, font = 3)
boxplot(Petal.Length ~ Species, data = iris,  col = colors,
        xaxt = "n")
axis(1, at = 1:3, labels = sp, font = 3)
boxplot(Petal.Width ~ Species,
        data = iris,
        col = colors,
        xaxt = "n")
axis(1, at = 1:3, labels = sp, font = 3)
par(mfrow = c(1,1))
```

---
## plot devices 

+ `plot()` opens a new graphic "device": a new window

--

+ `hist()`, `barplot()`, `boxplot()` also call for new devices

--

+ `points()` and `lines()` do not open new devices and need to be executed after `plot()` calls

--

+ new `plot()` calls reset the graphic device. 

--

+ `dev.off()` turns off the current plot device

---
## saving plots

+ to save plots in base R, new graphic devices must be called: `png()`, `pdf()`, etc- (check `capabilities()`)

--

+ basic recommended formats: __.png__ and __.tiff__ because they are not lossy (try not to use __.jpeg__)

--

+ `png()` calls for a new graphic device _different from the graphic window_

  + plot code

+ `dev.off()` to close the device and save. 

### you won't see the plot when you do that

---
## saving our plot

```{r, eval = F, echo = F}
colors <- c("#3B9AB2", "#EBCC2A", "#F21A00")
sp <- paste("I.", unique(iris$Species), sep = " ")
png(filename = "./figs/figure01.png")
par(mfrow = c(2, 2),
    mar = c(4, 3, 2, 2), 
    bty = 'l',
    las = 1)
boxplot(Sepal.Length ~ Species,
        data = iris,
        xlab = "",
        col = colors,
        xaxt = "n")
axis(1, at = 1:3, labels = sp, font = 3)
boxplot(Sepal.Width ~ Species, 
        data = iris, 
        xlab = "", 
        col = colors,
        xaxt = "n")
axis(1, at = 1:3, labels = sp, font = 3)
boxplot(Petal.Length ~ Species, data = iris,  col = colors,
        xaxt = "n")
axis(1, at = 1:3, labels = sp, font = 3)
boxplot(Petal.Width ~ Species,
        data = iris,
        col = colors,
        xaxt = "n")
axis(1, at = 1:3, labels = sp, font = 3)
dev.off()
```


---
class: middle
## data visualization has many don'ts

.pull-left[
```{r, echo = F}
library(lattice)
library(latticeExtra)
d <- read.table(text = ' x   y     z
t1   5   high
t1   2   low
t1   4   med
t2   8   high
t2   1   low
t2   3   med
t3  50   high
t3  12   med
t3  35   low', header = TRUE)
d$x <- factor(d$x)
d$z <- factor(d$z)
levels(d$z) <- c("high", "med", "low")

cloud(y ~ x + z, d,
      panel.3d.cloud = panel.3dbars,
      col.facet = d$z,
      xbase = 0.4,
      ybase = 0.4,
      scales = list(arrows = FALSE, col = 1),
      par.settings = list(axis.line = list(col = "transparent")),
      main = "my 3D plot")
```
]

--

.pull-right[
```{r, echo = F, message=F, warning=F, out.width=400}
library(ggplot2)
pal <- wesanderson::wes_palette(3, name = "Zissou1", type = "continuous")
p <- ggplot(data=d, aes(x=x, y=y, fill=z)) +
  geom_bar(stat="identity") +
  scale_fill_manual(values=pal[3:1])+
  theme_classic() + 
  theme(legend.text = element_text(size = 16),
        legend.title = element_text(size = 16))
p
```
]

---

## many

.pull-left[
```{r pie3d, echo=FALSE}
library(plotrix)
slices <- c(10, 12, 4, 16, 8) 
lbls <- c("US", "UK", "Australia", "Germany", "France")
pct <- round(slices / sum(slices) * 100)
lbls <- paste(lbls, pct) # add percents to labels 
lbls <- paste(lbls, "%", sep = "") # ad % to labels 
pie3D(slices,labels = lbls, explode = 0.1,
   main = "Pie Chart of Countries ")
```
]

--

.pull-right[

```{r, echo = F}
pie(slices,labels = lbls, col=rainbow(length(lbls)),
   main="Pie Chart of Countries")
```

]

---
## there's always a better option than a pie chart

.pull-left[

```{r barplot, echo=FALSE}
nomes <- c("US", "UK", "Australia", "Germany", "France")
names(pct) <- nomes
perc <- paste0(pct, "%")
par(mai=c(1,1,1,1))
barplot(pct, horiz=TRUE, las=1, xlim=c(0, max(pct)+3))
text(x=pct+1.2,
     y=c(0.7, 1.9, 3.1, 4.3, 5.5),
     labels=perc)
```

]

--

.pull-right[

```{r barplot2, echo=FALSE}
nomes <- c("US", "UK", "Australia", "Germany", "France")
perc <- paste0(sort(pct), "%")
names(pct) <- nomes
par(mai=c(1,1,1,1))
barplot(sort(pct), horiz=TRUE, las=1, xlim=c(0, max(pct)+3))
text(x=sort(pct)+1.2,
     y=c(0.7, 1.9, 3.1, 4.3, 5.5),
     labels=perc)
```

]

---
## barplots are not always very informative

```{r, echo = F}
set.seed(42)
d2 <- data.frame(name = letters[1:5],
                 value = sample(seq(4, 15), 5),
                 sd = c(1, 0.2, 3, 2, 4))

barplot(value ~ name, data = d2, las = 1, bty = 'l')
```

---
## better with error bars...

```{r, echo = F}
barplot(value ~ name, data = d2, las = 1,
        bty = 'l', ylim = c(0, 18))
arrows(x0 = c(0.66, 1.88, 3.07, 4.25, 5.48),
       y0 = d2$value + d2$sd,
       y1 = d2$value - d2$sd, angle = 90, length = 0.1, code = 3)
```

---
## but maybe don't even make a barplot

```{r, echo = F}
plot(x = 1:5, d2$value, las = 1, bty = 'l', ylim = c(0, 18),
     pch = 19, xaxt = 'n',
     xlab = "names", ylab = "value")
axis(1, at = 1:5, labels = d2$name)
arrows(x0 = 1:5,
       y0 = d2$value + d2$sd,
       y1 = d2$value - d2$sd,
       angle = 90, length = 0.05, code = 3)
```

---
##  ...or maybe don't even make a graph


```{r, echo = F}
d3 <- data.frame(media = c(6.7, 12.3),
                 sd = c(0.4, 0.98),
                 tratamento = c(1, 2))
plot(media ~ c(1,2), data = d3, las = 1, bty = 'l',
     xlab = "tratamento", pch = 19, xaxt = 'n',
     ylim = c(2, 15))
axis(1, at = 1:2, labels = c("1", "2"))
arrows(x0 = 1:2,
       y0 = d3$media + d3$sd,
       y1 = d3$media - d3$sd,
       angle = 90, length = 0.05, code = 3)
```

---
## make a table or say it in the text 

|Treatment| Effect|
|-----------|--------|
|1 |`r d3$media[1]` $\pm$ `r d3$sd[1]` | 
|2 |`r d3$media[2]` $\pm$ `r d3$sd[2]` | 


---
## some basic tips in general

+ only make plots when you really need to

--

+ don't spend more ink and colors than you need to

--

+ don't fool your reader (no y-axis tampering, no undue transformation)

--

+ show error measures


---
## some basic tips in R

--

+ use `las = 1` for your axis labels

--

+ use `bty = "l"` for your boxes

--

+ change to at least `pch = 19`

--

+ use `xlab` and `ylab`

--

+ save to png and pdf formats

---
## Statistical procedures: package __stats__

+ Linear regression: `lm()`
+ Analysis of variance: `anova()`, `aov()`
+ t-tests: `t.test()`
+ p-values correction: `p.adjust()`

__R TASKVIEWS__  [https://cran.r-project.org/web/views/](https://cran.r-project.org/web/views/)

---
class: center, middle
# ¡Thanks!

<center>
`r icons::icon_style(icons::fontawesome("paper-plane", style = "solid"), scale = 1, fill = "#A70000")` [andreasancheztapia@gmail.com](mailto:andreasancheztapia@gmail.com) 

`r icons::icon_style(icons::fontawesome("twitter", style = "brands"), scale = 1, fill = "#A70000")` [@SanchezTapiaA](https://twitter.com/SanchezTapiaA) 

`r icons::icon_style(icons::fontawesome("github", style = "brands"), scale = 1, fill = "#A70000")``r icons::icon_style(icons::fontawesome("gitlab", style = "brands"), scale = 1, fill = "#A70000")` `r icons::icon_style(icons::fontawesome("bitbucket", style = "brands"), scale = 1, fill = "#A70000")` [andreasancheztapia](http://github.com/andreasancheztapia) 
